---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# reapeR

<!-- badges: start -->
<!-- badges: end -->

`reapeR` is a developmental R package providing utilities for interacting with [David Talkin's *Ropust Epoch and Pitch EstimatoR* (REAPER) software](https://github.com/google/REAPER). REAPER provides precise pitch tracking and epoch estimation from speech data. It is a command line tool that outputs results in a raw text format. The main purpose of the `reapeR` library is to clear the potential hurdle of installing and working with such a command line tool. `reapeR` provides functionality for installing REAPER locally, for running the software (in bulk or on a single sound file), for loading in output files generated by REAPER using well-formatted R data structures, and for converting REAPER output into files that can be used with [Praat](https://www.fon.hum.uva.nl/praat/). 

This is a beta version of `reapeR`! The functions for interacting with REAPER are command line interfaces and as such, can be unstable across machines. An unquestionably better solution to these problems would be to use `Rcpp` or similar to directly integrate the REAPER source code into R without the unstable command line overhead, but this is somewhat above my pay grade (and building a command line interface was comparatively fast and painless). If you run into any problems with the package, I'm happy to hear about them.

## Installation

You can install the development version of `reapeR` from [GitHub](https://github.com/) with:

``` {r}
#| eval: false

# install.packages('devtools')
devtools::install_github('rpuggaardrode/reapeR')
```

This should also install REAPER along with the R package. If this works as intended, you should see a bunch of progress messages in the R console ending with the message `[100%] Built target reaper`. If this does *not* work for some reason, you can try again with the `reaper_install()` function.

```{r}
#| eval: false

reapeR::reaper_install()
```

If REAPER is *still* not installed, this could be because either [Git](https://github.com/git-guides/install-git) or [CMake](https://cmake.org/download/) are not available on your machine. They are required for the installation to work. Getting CMake set up right can be a bit of a hassle -- if you are on a Mac, the preferred way to install CMake is by running the following code from the terminal:

`brew install cmake`

## Running REAPER from R

If REAPER was successfully installed, you can now run the software from R using the `reaper()` function. In the simplest use case, you just pass a single argument with the name of a sound file.

```{r example}
library(reapeR)
reaper_out <- reaper('inst/extdata/1.wav')
```

The resulting object, `reaper_out`, is a `list` containing two elements called `pitch` and `epochs`. 

```{r}
class(reaper_out)
names(reaper_out)
```

`pitch` is a tibble with columns including information about the time of each analysis frame, a column with information about whether the frame is voiced, a column with the predicted *F*~0~, and a column with the file name.

```{r}
head(reaper_out$pitch)
```

The predicted pitch track looks like this (zooming into the middle of the file where there is voicing):

```{r}
plot(reaper_out$pitch$time, reaper_out$pitch$f0, type = 'l',
     xlab = 'Time (s)', ylab = 'F0 (Hz)',
     xlim = c(0.5, 1))
```

`epochs` is just a named vector of values telling you the locations of glottal closure instants:

```{r}
reaper_out$epochs
```

Here they are plotted along with the sound file in question:

```{r}
snd <- tuneR::readWave('inst/extdata/1.wav')
times <- seq(0, length(snd@left) / snd@samp.rate, 
             length.out = length(snd@left))
plot(times, snd@left, type = 'l', xlab = 'Time (s)', ylab = '', yaxt = 'n',
     xlim = c(0.65, 0.95))
abline(v = reaper_out$epochs[[1]], col='lightgrey')
```

You can use the same arguments with `reaper()` as when running the program from the command line (pitch floor and ceiling, analysis interval, etc). These arguments can be inspected by calling `?reaper` from the console. If you're only interested in the `pitch` data frame, set `output = 'pitch'`. If you just want epochs, set `output = 'epochs'`. If you can't get the REAPER installation to work with the R package but you already have REAPER in your system and want to control it with R, you can set the path to your REAPER executable with the `exePath` argument.

If you want to run REAPER on all files in a directory, you can do so with the `reaper_bulk()` function, where you pass the path to a directory instead of the path to a sound file:

```{r}
reaper_output <- reaper_bulk('inst/extdata')
```

`reaper_bulk()` has a few more interesting options. 

There's the argument `hirst2pass`, which can be set to `TRUE` to use the Hirst/De Looze two-pass procedure to estimate suitable pitch floor and ceiling values. In this case, REAPER is run with very liberal floor and ceiling values on the directory, and the resulting pitch is used to estimate more suitable values for a second pass.

Additionally, the `praat_output` argument can be set to `TRUE`, in which case the REAPER estimated pitch is also saved as `.Pitch` files that can be read in Praat (to a directory specified with `praat_output_dir`).

## Working with REAPER output

If you don't want to call REAPER from R or can't get the command line interface to work, the library also comes with several functions to work with REAPER output files.

`read_pitch_out()` will read an ASCII-formatted REAPER pitch output file (saved with `-f filename -a`) as a well-formatted tibble.

```{r}
pitchOut <- read_pitch_out('inst/extdata/pitch')
head(pitchOut)
```

`read_epochs_out()` will read an ASCII-formatted REAPER epochs ('pitchmarks') output file (saved with `-p filename -a`) as a vector.

```{r}
epochsOut <- read_epochs_out('inst/extdata/epochs')
epochsOut
```

If you're interested in working with this output in Praat, the `write_praat_pitch()` function will save a tibble with pitch output as a `.Pitch` file, and the `write_praat_epochs()` function will save a vector with epoch output as a `.PointProcess` file.

If you want to import REAPER-estimated pitch into an [EMU database](https://rpuggaardrode.github.io/emuintro/), you can convert a tibble with pitch output to SSFF files using the `reaper2ssff()` function.
